---
title: 'DPSS Capstone, Social Inequality: George Floyd Protests'
author: "Chanteria Milner"
date: '2021-08-03'
output: pdf_document
---

```{r setup, echo=FALSE, include=FALSE}
library(tidyverse)
library(sf)
library(ggplot2)
library(usmap)
library(readxl)
library(viridis)
library(formatR)
library(kableExtra)
library(reshape2)
library(stargazer)
knitr::opts_chunk$set(echo = TRUE, comment=NA, tidy.opts=list(width.cutoff=60),tidy=TRUE)

setwd("[INSERT WORKING DIRECTORY]")
```


## Task 1 - Setup

```{r setup2, include=FALSE, echo=FALSE, message=FALSE}
# Set up for census data
library(jsonlite)   # For reading raw JSON
library(tidycensus) # For downloading U.S. Census data
census_api_key("[INSERT API KEY]", install = TRUE, overwrite=TRUE)
readRenviron("~/.Renviron")
Sys.getenv("CENSUS_API_KEY")
```

```{r message=FALSE, warning=FALSE}
# reading in relevant data - base
protests <- read_csv("data/george-floyd-exports-june-22.csv")
us_counties <- st_read("data/tl_2016_us_county/tl_2016_us_county.shp")
```

```{r message=FALSE, warning=FALSE}
# getting gini index data
vote_2020 <- read_excel("data/vote2020.xlsx")
vars_acs_2019 <- load_variables(2019, "acs5")
gini_index <- get_acs( # note: 0 = no inequality, 1 = complete inequality
  geography = "county",
  variables = c("B19083_001"),
  year = 2019
)
gini_index <- gini_index %>%
  select(-variable)
```

```{r message = FALSE}
# getting parler data
parler_data <- read_csv("data/parler-videos-geocoded.csv")
```

## Task 2 - Generating Maps
### PART 1 - Sub-setting protest data for variables of interest

```{r}
protests_map <- protests %>%
  select(longitude, latitude, type_of_gathering, state, 
         escalation, police_altercation) %>%
  mutate(
    size = case_when(
    protests$size == "Large" ~ 2,
    protests$size == "Moderate" ~ 1,
    protests$size == "Small" ~ 0),
    type_of_gathering = ifelse(protests$type_of_gathering=="Protests", 0, 1))
```

### PART 2 - Mapping Data

```{r message=FALSE, warning=FALSE}
protests_map_trans <- usmap_transform(protests_map) # transform for usmaps

protests_map_trans <- # remove data Hawaii and Alaska
  protests_map_trans[protests_map_trans$state !="Hawaii" & 
                       protests_map_trans$state !="Alaska",]
```

#### 2a: Mapping general location of protests across the us

```{r echo=FALSE, out.width="80%", fig.align='center'}
plot_usmap(fill="darkolivegreen", alpha=0.35, color="white", exclude=c("AK", "HI")) +
  geom_point(data=protests_map_trans, aes(x=longitude.1, y=latitude.1),
             color = "darkblue", alpha = 0.25, size=2) +
  theme(
    panel.background = element_rect(color = "black", fill = "aliceblue"),
    plot.title = element_text(size=14),
    plot.subtitle = element_text(size=12),
    ) +
  labs (
    title = "Location of George Floyd Protests, Summer 2020",
    subtitle = "Source: Ipsos")
```

#### 2b: Mapping protest across the us by size

```{r echo=FALSE, out.width="80%", fig.align='center'}
plot_usmap(fill="darkolivegreen", alpha=0.35, color="white", exclude=c("AK", "HI")) +
  geom_point(data=na.omit(protests_map_trans), aes(x=longitude.1, y=latitude.1, 
                                                   size=size, color=factor(size)),
             alpha = 0.5) +
  scale_size_continuous(
    name = "Size of Protest",
    breaks = c(0,1,2),
    range = c(1, 3),
    labels = c("Small", "Moderate", "Large")
  ) +
  scale_color_manual (
    name = "Size of Protest",
    values = c("black", "blue", "firebrick2"),
    labels = c("Small", "Moderate", "Large"),
  ) +
  theme(
    panel.background = element_rect(color = "black", fill = "aliceblue"),
    plot.title = element_text(size=14),
    plot.subtitle = element_text(size=12),
    legend.text = element_text(size=10),
    legend.title = element_text(size=11),
    legend.background = element_rect(color="black")
    ) +
  labs(
    title = "Location of George Floyd Protests by Size, Summer 2020",
    subtitle = "Source: Ipsos",
    size = "Size of Protests"
  )
```

#### 2c: Mapping protest by if they escalated to violence

```{r echo=FALSE, out.width="80%", fig.align='center'}
plot_usmap(fill="darkolivegreen", alpha=0.35, color="white", exclude=c("AK", "HI")) +
  geom_point(data=protests_map_trans, 
             aes(x=longitude.1, y=latitude.1, color=factor(escalation),
                 shape=factor(escalation)
                 ), 
             size=2) +
  scale_color_manual (
    name = waiver(),
    values = c("darkblue", "red"),
    labels = c("Protests", "Protests that Escalated to Violence")
    ) +
  scale_shape_manual(
    name = waiver(),
    values = c(20, 17),
    labels = c("Protests", "Protests that Escalated to Violence")
  ) + 
  theme(
    panel.background = element_rect(color = "black", fill = "aliceblue"),
    legend.title = element_blank(),
    plot.title = element_text(size=14),
    plot.subtitle = element_text(size=12),
    legend.text = element_text(size=10),
    legend.background = element_rect(color="black")
    ) +
  labs(
    title = "George Floyd Protests by Escalation to Violence, Summer 2020",
    subtitle = "Source: Ipsos")
```

#### 2d: Mapping protest by if they had police violence

```{r echo=FALSE, out.width="90%", fig.align='center'}
protests_map_trans_p <- # remove if police violence not recorded
  protests_map_trans[protests_map_trans$police_altercation !="Not recorded",]

protests_map_trans_p <- protests_map_trans_p %>%
  mutate(police_altercation = case_when (
    police_altercation == "No" ~ "Protests",
    police_altercation == "Yes" ~ "Protests with Police Altercation"
  ))

plot_usmap(fill="darkolivegreen", alpha=0.35, color="white", exclude=c("AK", "HI")) +
  geom_point(data=protests_map_trans_p, aes(x=longitude.1, y=latitude.1, 
                                          color=factor(police_altercation), 
                                          shape=factor(police_altercation)),
             size=2, alpha=0.35) +
  facet_wrap(
    ~police_altercation
    )+
  scale_color_manual (
    name = waiver(),
    values = c("darkblue", "red"),
  ) +
  scale_shape_manual(
    name = waiver(),
    values = c(20, 17),
  ) +
  theme(
    panel.background = element_rect(color = "black", fill = "aliceblue"),
    legend.position = "none",
    strip.text.x = element_text(size = 10),
    plot.title = element_text(size=11),
    plot.subtitle = element_text(size=9)
    ) +
  labs(
    title = "George Floyd Protests by Police Altercation, Summer 2020",
    subtitle = "Source: Ipsos",
  )
```

## Task 3 - Running Regressions

```{r setup3, include=FALSE, warning=FALSE, message=FALSE}
# Set up - packages for creating tables from regression results
library(sjPlot)
library(sjmisc)
library(sjlabelled)
```

### PART 1 - Collapsing and generating variables
#### 1a: Retrieving county names for protest data

```{r}
protests_geom <- protests %>% # choosing variables of interest
  select(longitude, latitude, escalation) %>%
  mutate(escalation = ifelse(escalation=="No", 0, 1))

protests_geom <- st_as_sf( # generating correct geom cat for county data join
  protests_geom,
  coords = c("longitude", "latitude"),
  crs = 4326)

counties_sub <- us_counties %>% # subsetting county data set
  select(fips, STATEFP) 

counties_sub <- st_transform(counties_sub, 4326) # for merge with protest data

protests_geom <- st_join( # joining protest and county to get protest/county
  protests_geom,
  counties_sub,
  join = st_within
)
```

#### 1b: calculating num protests + escalations that escalated per County

```{r warning=FALSE, message=FALSE}
protests_geom$fips <- as.character(protests_geom$fips) # makes grouping by fips easier
protests_geom <- protests_geom %>%
  select(-STATEFP) # will add back later

protests_geom <- protests_geom %>%
  select(escalation, fips) %>% 
  add_count(fips) %>% # creates count variable of num protests that happened
  rename(num_protests = n) %>% # renames count variable
  select(escalation, fips, num_protests) %>%
  group_by(fips, num_protests) %>% 
  summarise(num_escalations = sum(escalation)) # creates count of escalations
```

### PART 2 - Joining new variables with county-level voting data
#### 2a: cleaning county name and fips code in vote data

```{r}
vote_2020 <- vote_2020 %>%
  rename(County = county_name,
         fips = county_fips)

vote_2020$fips <- as.character(vote_2020$fips) # makes joining easier
```

#### 2b: joining voting data with protest data + changing all na values

```{r}
vote_2020_protest <- left_join(vote_2020, protests_geom, by="fips")

vote_2020_protest$num_escalations[is.na(vote_2020_protest$num_escalations)] <- 0

vote_2020_protest <- vote_2020_protest %>% # add escalate binary variable
  mutate(escalated_b = ifelse(num_escalations == 0,0,1))

vote_2020_protest$num_protests[is.na(vote_2020_protest$num_protests)] <- 0

vote_2020_protest <- vote_2020_protest %>% # add if protest happened binary variable
  mutate(protest_b = ifelse(num_protests == 0,0,1))

vote_2020_protest <- vote_2020_protest %>% # don't need anymore
  select(-geometry)

# get statefp from counties dataset for fixed effects later
counties_sub$fips <- as.character(counties_sub$fips)
vote_2020_protest <- left_join(vote_2020_protest, counties_sub, by="fips")
```

### PART 3 - Running regressions on generated data
#### 3a - effect between protest occurrence and escalation on Trump vote share

```{r}
vote_2020_protest$esc_prot_interact <- # hard-code interaction term b/c NAs
  vote_2020_protest$escalated_b*vote_2020_protest$protest_b

reg_prot_viol_vote <- lm(votes_gop ~ protest_b + esc_prot_interact, 
                            data=vote_2020_protest) # looking at num votes

reg_prot_viol_vote2 <- lm(per_gop ~ protest_b + esc_prot_interact, 
                          data=vote_2020_protest) # looking at proportion of votes

# num gop votes
summary(reg_prot_viol_vote)

# proportion gop votes
summary(reg_prot_viol_vote2)
```

#### 3b - effect between protest counts and escalation on Trump vote share

```{r}
reg_prot_cnt_viol_vote <- lm(votes_gop ~ num_protests + escalated_b*num_protests, 
                             data=vote_2020_protest) # num gop votes
reg_prot_cnt_viol_vote2 <- lm(per_gop ~ num_protests + escalated_b*num_protests, 
                              data=vote_2020_protest) # proportion gop votes

# num gop votes
summary(reg_prot_cnt_viol_vote)

# proportion gop votes
summary(reg_prot_cnt_viol_vote2)
```

## TASK 4 - Fixed Effects
### PART 1 - effect between protest occurrence and escalation on Trump vote share

```{r}
fe_prot_viol_vote <- lm(
  per_gop ~ protest_b + esc_prot_interact + factor(STATEFP),
  data=vote_2020_protest)
summary(fe_prot_viol_vote)
```

### Part 2 - effect between protest counts and escalation on Trump vote share

```{r}
prot_cnt_viol_vote_fe <- lm(
  per_gop ~ num_protests + escalated_b*num_protests + factor(STATEFP),
  data=vote_2020_protest)
summary(prot_cnt_viol_vote_fe)
```

## EXTRA 1 - income inequality data below
### PART 1: joining gini index data with us counties data to get fips codes

```{r warning=FALSE}
counties_fips_geo <- us_counties %>% # to make join easier
  select(GEOID, fips)

counties_fips_geo <- st_transform(counties_fips_geo, 4326)

gini_geom <- left_join(gini_index, counties_fips_geo, by = "GEOID")

gini_geom <- gini_geom %>%
  rename(coefficient = estimate) %>%
  select(-moe)
```

### PART 2: joining protest data with gini data to get index for each protest county + map
#### 2a - joining gini and protest data

```{r warning=FALSE}
protests_fips <- protests %>% # creating additional lat/long for later mapping
  select(longitude, latitude, size:state_deployed_national_guard) %>%
  mutate(longitude.1 = longitude,
         latitude.1 = latitude)

protests_fips <- st_as_sf( # to join with county data to get fips
  protests_fips,
  coords = c("longitude.1", "latitude.1"),
  crs = 4326
)

counties_fips <- counties_fips_geo %>%
  select(-GEOID)

protests_fips <- st_join( # joining protest and county to get protest/county
  protests_fips,
  counties_fips,
  join = st_within
)

gini_fips <- gini_geom %>% # get just gini coefs and fips
  select(fips, coefficient)

protests_fips_gini <- left_join(protests_fips, gini_fips, by = "fips")

protests_fips_gini <- rename(protests_fips_gini, gini_coef = coefficient)

protests_fips_gini_trans <- usmap_transform(protests_fips_gini)
```

#### 2b - mapping protests by gini coef

```{r echo=FALSE, out.width="80%", fig.align='center'}
plot_usmap(fill="darkolivegreen", alpha=0.35, color="white") +
  geom_point(data=protests_fips_gini_trans, aes(x=longitude.1, y=latitude.1, color=gini_coef),
             size=2) +
  theme(
    panel.background = element_rect(color = "black", fill = "aliceblue"),
    plot.title = element_text(size=12),
    plot.subtitle = element_text(size=10),
  ) +
  scale_color_viridis(
    option = "magma"
    )+
  labs (
    title = "Location of George Floyd Protests by Gini Coefficient, Summer 2020",
    subtitle = "Sources: Ipsos and American Community Survey 2019",
    color="Gini Coef.")
```

### PART 3: add ginis to data on unique fips + protests
#### 3a - add fips to protests_geom data to get protests counts + coef per fip

```{r}
gini_fips$fips <- as.character(gini_fips$fips)

pro_fips_uniq <- left_join(protests_geom, gini_fips, by="fips")

pro_fips_uniq <- pro_fips_uniq %>%
  rename(gini_coef = coefficient)

# 3b: join with votes
vote_2020_gini <- left_join(vote_2020, gini_fips, by="fips") # to get coefs for each fip

vote_2020_gini <- left_join(vote_2020_gini, pro_fips_uniq, by="fips") # to get protest data

vote_2020_gini <- vote_2020_gini %>% # drop unnecessary/duplicate variables
  select(-geometry, -gini_coef)
```


```{r echo=FALSE}
# fix NAs
vote_2020_gini$num_protests[is.na(vote_2020_gini$num_protests)] <- 0
vote_2020_gini$num_escalations[is.na(vote_2020_gini$num_escalations)] <- 0

vote_2020_gini <- vote_2020_gini%>% # add binary variables
  rename(gini_coef = coefficient) %>%
  mutate(
    escalate_b = ifelse(num_escalations==0,0,1),
    protest_b = ifelse(num_protests==0,0,1)
  )
```

### PART 4: Regressions
```{r}
reg_num_prot_gini <- lm(num_protests ~ gini_coef, data=vote_2020_gini)
summary(reg_num_prot_gini) 
```

## EXTRA 2 - parler data below
```{r setup4, echo=FALSE}
library(lubridate)
election_2020 <- as.Date("2020-11-03")
jan_sixth <- as.Date("2021-01-06")
prot_start <- as.Date("2020-05-28")
prot_end <- as.Date("2020-06-25")
```

### PART 1: Mapping parler data by protests
#### 1a - isolate us parler data by protest location

```{r warning=FALSE}
parler_data_geom <- parler_data %>%  # extra lat long vars for future mapping
  mutate(longitude.1 = Longitude,
         latitude.1 = Latitude)

parler_data_geom <- st_as_sf( # gen geom for join with counties data
  parler_data_geom,
  coords=c("longitude.1", "latitude.1"),
  crs = 4326
)

protests_fips_2 <- as.data.frame(protests_fips$fips) # get protest fips codes

protests_fips_2<- rename(protests_fips_2, "fips" = "protests_fips$fips")


parler_geom_us <- st_join( # add all us fips to parler data
  parler_data_geom,
  counties_fips,
  join = st_within
)

parler_geom_us <- parler_geom_us[!is.na(parler_geom_us$fips),] # remove non us locations

parler_geom_prot <- parler_geom_us %>% # only keep parler data where protests happened
  semi_join(protests_fips_2)


parler_geom_prot <- usmap_transform(parler_geom_prot) # transform for mapping
```

#### 1b - map general parler data before and after protests start

```{r echo=FALSE, out.width="90%", fig.align='center'}
parler_geom_prot <- parler_geom_prot[-c(3472),] # remove random outlier point

parler_geom_prot <- parler_geom_prot %>% # create variable for activity pre/post protests
  mutate(post_protest = ifelse(Timestamp < prot_start, 0, 1))

parler_geom_prot$post_protest2 <-  # labels for facet wrap map
  factor(parler_geom_prot$post_protest, labels 
         = c("Before May 28, 2020", "After May 28, 2020"))


plot_usmap(fill="darkolivegreen", alpha=0.35, color="white") +
  geom_point(data=parler_geom_prot, aes(x=Longitude.1, y=Latitude.1),
             color = "red4", alpha = 0.25, size=2) +
  theme(
    panel.background = element_rect(color = "black", fill = "aliceblue"),
    plot.title = element_text(size=14),
    plot.subtitle = element_text(size=12)
  ) +
  facet_wrap(
    .~post_protest2,
  ) +
  labs (
    title = "Parler Activity Before and After Protests Start")
```

